---
title: "Automatic Replacement Tools"
output: 
  rmarkdown::html_vignette:
    df_print: kable
vignette: >
  %\VignetteIndexEntry{Automatic Replacement Tools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  rows.print = 6
)
library(pidpos)

.tmp <- tempdir()
pidpos:::set_model_folder(.tmp)
options(pidpos_download_approved = TRUE)
{
  model <- "english-ewt-ud-2.5-191206.udpipe"
  to <- file.path(.tmp, model)
  url <- file.path("https://raw.githubusercontent.com/jwijffels/udpipe.models.ud.2.5/master/inst/udpipe-ud-2.5-191206/", model)
  curl::curl_download(url, to, mode = "wb")
}
```

This package is designed to support the identification and redaction of personally identifiable data (PID) within textual datasets.

The `report_to_redaction_rules()` function generates a `replacement_rules` table from the output of `pidpos()`. This table specifies candidate values for redaction and provides a `To` column in which users can define replacement values.

In many cases, users may wish to manually populate the To column (for example, replacing names with consistent pseudonyms). However, when a dataset contains a large volume of PID, manual specification may be impractical. In such cases, the `auto_replace()` utility can be used to automatically generate replacement values.

`auto_replace()` operates on a `replacement_rules` table and encodes the `To` column according to a user-defined replacement function. The package provides three built-in replacement strategies:

- `make_hashing_replacement()` — hashes values using a key and salt, producing deterministic and reproducible replacements.
- `make_random_replacement()` — generates random replacements from a defined character space.
- `make_replacement_function()` - [Experimental] convert a function for generating a random string into a compatible function.

---

# Basic Workflow

We begin by generating a PID report and converting it into redaction rules:

```{r, results="hide", message=FALSE, eval=F}
library(pidpos)

report <- pidpos(the_one_in_massapequa)
replacement_rules <- report_to_redaction_rules(report)
replacement_rules
```

```{r, echo=FALSE, message=FALSE}
report <- pidpos(the_one_in_massapequa)
replacement_rules <- report_to_redaction_rules(report)

head(replacement_rules, 5)
```

To automatically populate the `To` column, we first initialise a replacement function.  
In this example, we use `make_random_replacement()` to generate five-character replacements drawn from upper-case letters:

```{r}
replacement_func <- make_random_replacement(
  replacement_size = 5,
  replacement_space = LETTERS
)
```

We then apply this function using `auto_replace()`:

```{r, results="hide"}
set.seed(101)
updated_replacement_rules <- auto_replace(
  replacement_rules,
  replacement_func
)

updated_replacement_rules
```

```{r, echo=FALSE}
head(updated_replacement_rules, 5)
```

The resulting `updated_replacement_rules`  can then be used alongside the original data in the `redact()` function to adjust the data:

```{r}
redact(
  head(the_one_in_massapequa, 5),
  updated_replacement_rules
)
```

If the quantity of text being redacted is large, and documents are regularly repeated, the user may wish 
to parse the replacement rules into a caching redaction function:

``` {r}
cached_redacter <- parse_redacter(updated_replacement_rules, with_cache = T)
cached_redacter
```

This new function has a `memoization` layer built in, so that if the same document is presented - replacements are called from memory.  This may speed up data processing if the same passage of text is presented multiple times, but comes at the cost of memory.  The `cached_redact_function` can be used in `redact()` in the same way:

``` {r}
redacted_docs <- redact(
  the_one_in_massapequa,
  cached_redacter
)
cached_redacter
```

And its representation tracks the number of unique documents stored.

# User defined functions

Depending on the situation, users may wish to implement their own replacement functions.  
Where these are purely deterministic a single-argument function can be passed into `auto_replace()` to generate new values:

``` {r}
simple_replacement <- function(vec) {
  gsub(".*", "XXX", vec)
}

auto_replace(
  head(replacement_rules),
  simple_replacement
)
```

In addition, the user may wish to generate their own pseudo-random replacements.  
This can be done by wrapping the desired behaviour in `make_replacement_function()` with 
the added requirement of defining the maximum number of random states that can be generated (e.g. if we used a toy example of the number "00" to "99" there would be 100 max random values).  We implement this as:

``` {r}
numeric_replacement <- function() {
  paste(sample(0:9, 2, T), collapse = "")
}

numeric_replacement <- make_replacement_function(numeric_replacement, 100)

numeric_replacement
```
This function can then be used in `auto_replace()` in the same way as the deterministic function:

``` {r}
auto_replace(
  head(replacement_rules),
  numeric_replacement
)
```

With the added benefit the functional representation tracks how many of the allowed values have been taken:

``` {r}
numeric_replacement
```
With utilities supplied to retrieve the underlying key-value pairs (see `get_replacement_cache()`, `key_lookup()` and `value_lookup()`):

``` {r}
get_replacement_cache(numeric_replacement)
```

---

# Summary

The automatic replacement tools allow users to:

- Rapidly generate redaction mappings for large datasets
- Choose between deterministic or randomised replacement strategies
- Convert replacement rules into reusable redaction functions

This approach separates PID identification from transformation, enabling reproducible and auditable redaction workflows.
