---
title: 'pidpos: An R package for the detection of personally identifiable data'
bibliography: ../JOSS_utilities/paper.bib
output: 
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{pidpos: An R package for the detection of personally identifiable data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

``` {r setup, echo=F}
knitr::opts_chunk$set(collapse = T, comment = "#>", warning = F)
options(tibble.print_min = 4, tibble.print_max = 4)
```

# Summary

The `pidpos` package aids in identifying personal identifiability risks in datasets. Using part-of-speech (POS) tagging, it extracts proper nouns from text fields, reducing the complexity of the review process and enabling faster human oversight. The package also provides tools for designing and implementing a redaction workflow.

# Statement of need  

Data collection and analysis has grown enormously in scale and scope, prompting international legislation to protect individuals' rights over their own data [@GDPR]. This has heightened awareness of the responsibilities of data controllers [@ICODataController] and the risks posed by large datasets [@clarke2016big]. A central concern is personal identifiability — the ability to directly or indirectly identify an individual from a dataset [@finck2020they] — with breaches carrying significant reputational and financial consequences [@ICOWhatIf].

For small, structured datasets, manual inspection can identify personally identifiable data (PID) with reasonable effort. In large datasets, however, PID embedded within free-text fields or appearing rarely in a variable can easily be missed. Existing R packages such as PII [@pii] address this through pattern matching, which risks missing edge cases.

`pidpos` takes a different approach. Building on *part-of-speech* tagging (by default the udpipe framework [@straka2016udpipe; @udpipecran], with the ability to use a custom tagging engine) it extracts all proper nouns from a dataset, deliberately accepting a higher false positive rate,  and implementing tools to aid human review rather than attempting full automation. This makes it robust to the edge cases that pattern-matching approaches can miss.

# In practice

To install the current version of `pidpos` package, use the following code:

``` {r install.1, eval=F}
# install.packages("pak")
pak::pkg_install("Stat-Cook/pidpos")
```


The intended workflow breaks down into three stages: 

1. Detection of PID risks via `pidpos()`
2. Preparation of redaction rules via `report_to_redaction_rules()` and `auto_replace()`
3. Redaction of the original data via `redact()`

To illustrate this, we include a subset of the `friends` package data set:

``` {r load, warning = F, message = F, results="hide"}
library(pidpos)
example_data <- head(the_one_in_massapequa, 20)
example_data
```

``` {r dataset, warning = F, message = F, echo=F}
example_data
```

First, generate a PID report:

``` {r}
report <- pidpos(example_data)
head(report)
```

The report lists all detected proper nouns alongside their source variable and position. By default, `pidpos()` uses the `udpipe` framework for POS tagging, but the package is designed to support alternative taggers. A ready-made script for using spaCy is included, and users may supply a custom tagging function, allowing the package to be integrated into existing NLP pipelines. Further details are provided in  [Custom Functions](https://stat-cook.github.io/pidpos/articles/custom-functions.html).  

Should the user wish to not only identify, but redact the data, the  report can be 
converted into redaction rules and apply replacements:

``` {r}
raw_rules <- report_to_redaction_rules(report)
replacement_func <- make_random_replacement()
prepared_replacements <- auto_replace(raw_rules, replacement_func)
head(prepared_replacements)
```

Users may define replacement values manually or use the built-in automatic replacement tools, which include options such as random replacement and encryption. Full documentation of the available replacement strategies is provided in [Automatic Replacement Tools](https://stat-cook.github.io/pidpos/articles/auto-replacement.html). Finally, apply the rules to produce a redacted dataset:

``` {r}
redacted_data <- redact(example_data, prepared_replacements)
head(redacted_data)
```

# Multiple file API

When a project involves multiple files, three additional functions support batch processing:

* `report_on_folder()` to generate PID reports 
* `get_distinct_redaction_rules()` to combine the distinct reports into a single set of raw redactions.
* `redact_at_folder()` to produce redacted copies of the data.

# Current applications

The `pidpos` package was developed for applications in the NuRS and AmReS research projects
which aim to extract and analyse retrospective operational data from NHS Trusts to understand
staff retention and patient safety.

# Contributions

The package was designed by RC, MA and SJ. Implementation was done by RC. Quality assurance was 
done by MA. Documentation was written by RC. Funding for the work was won by RC and SJ.

# Acknowledgements

The development of `pidpos` was part of the NuRS and AmReS projects funded by the
Health Foundation.

# References



